% =============================================================
% Short methods note. Honest, narrow, clean.
% Build: pdflatex main && pdflatex main
% =============================================================
\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,mathtools}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage[colorlinks=true,allcolors=blue]{hyperref}
\usepackage{microtype}

\newcommand{\R}{\mathbb{R}}
\newcommand{\jvp}{\textsc{jvp}}
\newcommand{\fd}{\textsc{fd}}

\title{\vspace{-2em}Forward-Mode Differential Geometry of Generative Manifolds:\\
Exact Higher-Order Pushforward and the Failure of Finite Differences}
\author{Jihong An \quad\texttt{ether9073@gmail.com}}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
We describe an exact, constant-memory instrument for the higher-order
differential geometry of a trained image generator's latent-to-image map,
built on forward-mode automatic differentiation (\jvp). A single forward pass
yields the exact pushforward (first-order Jacobian--vector product); composing
two forward passes yields the exact second-order pushforward (curvature term),
and the construction extends to arbitrary order at constant memory. We then make
a simple but practically important point precise: \emph{finite differences
cannot substitute for this instrument}. On a toy map with analytically known
derivatives, central finite differences (\fd) approximate the second-order term
with at best $\sim$$10^{-8}$ relative error in double precision---$3.5\times10^{8}$
times worse than the exact composed \jvp---and diverge under catastrophic
cancellation as the step shrinks; on real generators in single precision the
\fd{} second-order estimate never approaches the \jvp{} value at any step size.
As a demonstration of what the instrument reveals, we report that per-layer
curvature intensity decays monotonically across the synthesis layers of a
StyleGAN generator (8/8 attributes, Spearman $\in[-0.96,-0.80]$), i.e.\ curvature
concentrates in the coarse early layers. We deliberately make no claim that this
curvature signal is an actionable predictor or controller of edit damage; a
companion analysis finds that it is not, and we treat the instrument and the
finite-difference fact as the contribution.
\end{abstract}

\section{Introduction}
Let $G:\R^{d}\to\R^{3HW}$ be a trained generator mapping a latent code to an
image, and let $\gamma(\alpha)=z_0+\alpha\,u$ be a straight latent path along a
direction $u$. The local geometry of the image manifold along this path is
governed by the pushforwards
\[
\dot I(\alpha)=\frac{\partial}{\partial\alpha}G(\gamma(\alpha)),
\qquad
\ddot I(\alpha)=\frac{\partial^2}{\partial\alpha^2}G(\gamma(\alpha)),
\]
the first being the image-space velocity and the second the curvature
(acceleration) term. These are standard objects, but computing $\ddot I$ for a
generator that outputs millions of pixels is usually approached either by
finite differencing $G$, which is inaccurate (Section~\ref{sec:fd}), or by
materializing Jacobians via reverse-mode automatic differentiation, which is
memory-prohibitive at output dimension $3HW$. We instead use forward-mode
automatic differentiation, which gives $\dot I$ and $\ddot I$ \emph{exactly} in
one and two forward passes respectively, at the memory cost of a single forward
pass.

\section{Method: exact higher-order pushforward via composed \jvp}
Forward-mode AD evaluates, for a map $f$ and a tangent $v$, the
Jacobian--vector product $\partial f(x)[v]$ alongside $f(x)$ in one pass without
forming the Jacobian. Taking $f=G$, $x=\gamma(\alpha)$, $v=u$ gives the exact
first-order pushforward $\dot I(\alpha)$. The second-order term is the
directional derivative of the first, so composing forward mode with itself,
\[
\ddot I(\alpha)
=\frac{\partial}{\partial\alpha}\,\big[\partial G(\gamma(\alpha))[u]\big][u],
\]
is again exact and is evaluated by nesting two \jvp{} calls
(\texttt{torch.func.jvp} of a function that itself returns a \jvp). Memory stays
at one forward pass because no Jacobian is stored; the same nesting gives any
order. The instrument is generator-agnostic: we ran it unmodified on StyleGAN1,
StyleGAN2, and a latent-diffusion model.

\section{Finite differences cannot recover the curvature term}
\label{sec:fd}
A natural objection is that one could just finite-difference $G$. We show this
fails, with an analytic referee so that \jvp{} is not judged against itself.

\paragraph{Toy map with known derivatives.}
Let $f(\alpha)=\big[e^{0.7\alpha},\,\sin 3\alpha,\,(\alpha+0.4)^3,\,\tanh
2\alpha,\,\alpha^2\cos\alpha\big]$, whose first and second derivatives are known
in closed form. At $\alpha=0.3$ in double precision (finite differences' best
case), Table~\ref{tab:toy} shows the central-\fd{} second-order relative error
is U-shaped---truncation error at large step, catastrophic cancellation at small
step---bottoming at $8.6\times10^{-9}$ near $\varepsilon=10^{-4}$ and diverging
to $0.29$ by $\varepsilon=10^{-8}$. The composed \jvp{} attains
$2.5\times10^{-17}$ (machine precision), i.e.\ \fd's best is $3.5\times10^{8}$
times worse. No step size lets \fd{} recover the curvature.

\begin{table}[h]\centering
\begin{tabular}{rcc}
\toprule
$\varepsilon$ & central \fd{} 2nd-order rel.\ err.\ & \\
\midrule
$10^{-1}$ & $7.9\times10^{-3}$ & (truncation) \\
$10^{-2}$ & $8.0\times10^{-5}$ & \\
$10^{-4}$ & $8.6\times10^{-9}$ & (best) \\
$10^{-6}$ & $3.0\times10^{-5}$ & \\
$10^{-8}$ & $2.9\times10^{-1}$ & (cancellation) \\
\midrule
composed \jvp & $2.5\times10^{-17}$ & (exact) \\
\bottomrule
\end{tabular}
\caption{Second-order accuracy on the analytic toy map at $\alpha=0.3$,
double precision. Finite differences never approach \jvp.}
\label{tab:toy}
\end{table}

\paragraph{Real generator.}
On a StyleGAN bedroom generator in single precision, the \fd{} second-order
estimate's relative error versus \jvp{} rises monotonically as the step shrinks
across all tested attributes (e.g.\ for \texttt{view}, from $2.4$ at
$\varepsilon=10^{-1}$ to $561$ at $\varepsilon=10^{-5}$; third-order errors reach
$\sim$$10^{7}$), with no step recovering the curvature. The single-precision
arithmetic that generators use makes the cancellation regime far more severe
than the toy fp64 case.

\section{What the instrument reveals: per-layer curvature structure}
As a brief illustration, we apply the instrument layer-by-layer: restricting the
tangent to each synthesis layer in turn and measuring the mean curvature
intensity. Across the $8$ annotated StyleGAN-bedroom attributes, curvature
intensity decays monotonically from the coarse early layers to the fine late
layers ($8/8$ attributes, Spearman$(\text{layer},\text{intensity})\in[-0.96,
-0.80]$): curvature is concentrated where coarse structure is set. This is a
descriptive observation enabled by the exact instrument; we do not build a claim
on top of it here.

\section{Limitations and scope}
This is a tools-and-numerics note. The instrument is a thin, exact wrapper around
forward-mode AD; its value is in making exact higher-order pushforward routine
and in the precise finite-difference comparison, not in algorithmic novelty. We
explicitly do \emph{not} claim that the curvature signal predicts or controls
perceptual/identity edit damage: a separate, rigorous analysis (across
architectures, encoders, identity metrics, and $>$$2000$ matched edit pairs)
finds that, after controlling for edit magnitude, curvature carries at most a
weak, baseline-definition-dependent residual and is not an actionable
per-direction edit-risk signal. We report that negative result honestly
elsewhere and keep this note to what is clean: an exact instrument and the fact
that finite differences cannot replace it.

\end{document}
