% =====================================================
% 02_related.tex — short related work, ~0.75 page
% =====================================================

\paragraph{GAN interpretation and attribute editing.}
\textbf{Unit-level / segmentation-based dissection}
\cite{Bau2019GANDissection} probes individual generator units via
attached segmentation classifiers. \textbf{Boundary-based editing}
\cite{Shen2020InterFaceGAN, Yang2020HiGAN} learns linear hyperplanes
in $W^+$ for semantic attributes; we use the resulting boundaries as
inputs to our analysis. \textbf{Unsupervised discovery} approaches:
PCA on activations~\cite{Harkonen2020GANSpace}, closed-form
factorisation of the affine layer~\cite{Shen2021SeFa}, channel-wise
style manipulation~\cite{Wu2021StyleSpace}, contrastive direction
learning~\cite{Yuksel2021LatentCLR, Ren2022DisCo}, and
deformation-field editing~\cite{Pan2023DragGAN}. We compare against
all of these on attribute rediscovery and editing-task head-to-head
(§\ref{sec:experiments}).

\paragraph{Diffusion-model interpretation.}
The diffusion analogue of $W$-space is the U-Net bottleneck h-space
introduced by Asyrp~\cite{Kwon2023Asyrp}. Park et
al.~\cite{Park2023Riemannian} constructs a first-order pullback
Riemannian metric on h-space via Jacobian SVD;
Haas et al.~\cite{Haas2024Hspace} adds spectral analysis of the
denoiser Jacobian via forward-mode JVPs --- methodologically the
closest \emph{computational} precedent to our approach.
Diepeveen et al.~\cite{Diepeveen2025ScoreBasedPullback} and Saito \&
Matsubara~\cite{Saito2025Tangential} formalise pullback / Jacobian
Riemannian metrics for diffusion;
Karczewski et al.~\cite{Karczewski2025Spacetime} extends this to a
Fisher-Rao metric on the latent spacetime.
\textbf{Most relevant of all is Lobashev et al.~\cite{Lobashev2025HessianLatent},
who measure the Hessian of the Fisher information metric on the
\emph{latent} space of Stable Diffusion} --- the closest existing work
to our second-order analysis.

\emph{Our delta over these works.} Park, Diepeveen, Saito, and
Karczewski compute \emph{first-order} pullback metrics, not
second-order curvature. Haas computes a first-order Jacobian
spectrum (no Hessian). Lobashev computes a Hessian, but of the
\emph{Fisher information metric on latents} --- a different object
from our \emph{second fundamental form of the rendered image
manifold}. Our framework operates on the image manifold extrinsic
curvature, connects it to disentanglement / composition /
layer-localisation, and applies across StyleGAN1, StyleGAN2, and
Stable Diffusion.

Prompt-side editing methods --- Prompt-to-Prompt~\cite{Hertz2023P2P},
null-text inversion~\cite{Mokady2023NullText},
SEGA~\cite{Brack2023SEGA}, Concept
Sliders~\cite{Gandikota2024ConceptSliders},
FluxSpace~\cite{Dalva2025FluxSpace} --- operate in attention or
LoRA space rather than on the image manifold geometry.
DAAM~\cite{Tang2023DAAM} provides cross-attention saliency at the
prompt level; we compare our gradient saliency to DAAM directly in
\S\ref{sec:exp-baselines-results}.
NoiseCLR~\cite{Dalva2024NoiseCLR} extends contrastive direction
discovery to diffusion noise space; we add it to the head-to-head.

\paragraph{Score-network geometry of diffusion.}
Stanczuk et al.~\cite{Stanczuk2024IntrinsicDim} measure rank
deficiency of the score Jacobian to recover the intrinsic data
dimension. Higher-order ODE solvers
(GENIE~\cite{Dockhorn2022GENIE}) use JVPs through the U-Net for
better numerical accuracy. Our use of forward-mode AD through the
DDIM sampling chain extends the GENIE recipe to second-order
geometric measurement of the rendered image, complementing the
score-side analyses of~\cite{Stanczuk2024IntrinsicDim}.

\paragraph{Classical saliency, image attribution, and Grad-CAM.}
CAM~\cite{Zhou2016CAM} and Grad-CAM~\cite{Selvaraju2017GradCAM}
attribute classifier decisions to spatial regions via either pooled
weights or class-score gradients. Our pushforward saliency parallels
Grad-CAM in spirit but operates on the generator alone, with no
classifier --- the source of curvature is the model itself, not a
recognition head. As §\ref{sec:experiments} shows, CLIP-Grad-CAM
saliency and our JVP saliency are cross-domain near-orthogonal
($|r| < 0.11$ on bedroom, $< 0.02$ on FFHQ), establishing them as
complementary rather than competing measurements.

\paragraph{Differential geometry of neural-network manifolds.}
Natural gradient~\cite{Amari1998NaturalGradient} and geometric deep
learning~\cite{Bronstein2021GeometricDL} formalise the ambient
geometry of \emph{parameter} space and \emph{input} domains. Our
setting is different: we study the geometry of the manifold of
\emph{outputs} learned by a fixed generator. Zhu et
al.~\cite{Zhu2016ManipulationOnManifold} introduced the "natural
image manifold" formulation; we provide a tractable second-order
estimator on this manifold and connect curvature to disentanglement
and compositional editing.

\paragraph{Forward-mode autodifferentiation.}
The Pearlmutter trick~\cite{Pearlmutter1994HVP} for Hessian-vector
products underlies modern second-order optimisation
\cite{Martens2015KFAC}; PyTorch \texttt{torch.func.jvp} and
\texttt{jax.jvp} provide composable forward AD. Cobb et
al.~\cite{Cobb2025FoMoH} recently introduced FoMoH, a hyper-dual-number
implementation that returns first-order tangent and second-order
curvature in a single forward pass --- a strict generalisation of the
composed-JVP construction we use here.
We compose JVP twice to obtain the per-pixel second derivative in
one pass (§4), which is $10^3$-$10^4 \times$ faster than per-pixel
reverse mode for the high-dimensional-output regime studied here
(empirically verified, Track 11).

\paragraph{Encoders, inversion, and chart consistency.}
pSp~\cite{Richardson2021pSp}, e4e~\cite{Tov2021e4e}, and the
survey~\cite{Xia2022InversionSurvey} establish reconstruction-error
as the de facto encoder-quality metric. We argue (§5.5) that
\emph{first-order} chart consistency --- agreement of saliency
derivatives at the encoded code with those at the ground-truth code
--- is a strictly stronger signal: on bedroom, the saliency
correlation rises 45$\times$ while recon MSE only falls 17\% across
40k iterations of synthetic-supervision training.

A full survey of $\sim$60 additional references --- spanning
StyleGAN-NADA, EditGAN~\cite{Ling2021EditGAN}, NoiseCLR, LELSD,
WarpedGANSpace, Concept Algebra, LEDITS, and Hessian-Penalty --- is
in the supplementary material.
