% =====================================================
% 06_application.tex — applications of the framework
% =====================================================

\subsection{Curvature-based pre-selection of editing directions}
\label{sec:app-curvature-preselect}

The C2 ratio $\bar\rho_a$ from §\ref{sec:exp-c1} ranks attributes
\emph{by what kind of edit they perform} --- texture-only edits sit at
$\bar\rho \approx 1$, topology-changing edits at $\bar\rho \gtrsim 20$.
Two practical uses follow:

\paragraph{(a) Curvature as an editing-quality filter.}
Given a candidate pool of $N$ direction discoveries
(GANSpace, SeFa, LatentCLR, DisCo, random), filter to the
curvature-low subset $\{v : \bar\rho_v < \tau\}$ if the goal is
clean, identity-preserving edits, or the curvature-high subset if
the goal is intentional structural change. Track 2
(experiments/baselines/run\_editing\_head\_to\_head.py) implements
this on $N = 1000$ FFHQ test latents and reports paired ID-cos /
target-Δ-attribute / LPIPS at the StyleSpace equal-strength matching
\cite{Wu2021StyleSpace}. \todo[Result numbers fill here once Track 2
completes (ETA after Wave 1 queue).]

\paragraph{(b) Predicting compositional editing failure.}
Using the mixed-Hessian predictor $P(a, b)$ from §\ref{sec:exp-c4},
we ask: given two attribute boundaries $a, b$ in a domain we have
not yet tested compositionally, can we predict ahead of time
whether they will compose linearly or interfere? The honest answer
from §\ref{sec:exp-c4} is "yes, *between* curvature regimes; no,
*within*". A 28-pair bedroom train / 10-pair FFHQ test
hold-out gives a single regime-threshold classifier that
identifies structural-vs-textural pair behaviour at
$\sim 80\%$ accuracy.

\subsection{Unsupervised attribute discovery + CLIP labelling}
\label{sec:app-c6-pipeline}

The C6 pipeline (§\ref{sec:exp-c6}, code in
\texttt{experiments/domains/ffhq/run\_discovery\_c6.py}) yields
candidate attribute boundaries without any human-labelled training
data: sample $N$ random unit directions on per-layer W$^+$ spheres,
cluster, label cluster centroids with CLIP. We exploit this in two
ways:

\paragraph{Proposing unlabelled directions.}
On bedroom, the C6 clusters at top-3 CLIP labels recovered 5/8 HiGAN
boundaries (precision 1.00). The 3 missed attributes
(\texttt{wood}, \texttt{indoor\_lighting}, \texttt{cluttered\_space})
do not appear in the top-3 of any K=8 cluster, but the corresponding
phrases are in the CLIP vocabulary — \emph{the pipeline tells us
which attributes the generator's latent geometry doesn't natively
stratify}. We propose this as a diagnostic for attribute-taxonomy
adequacy.

\paragraph{Recall scaling with N (Track 12).}
Bedroom recall vs $N \in \{128, 192, 256, 384, 512\}$ tells us how
much sampling is sufficient. \todo[fill once Track 12 completes].

\subsection{The framework as a cross-method evaluator}
\label{sec:app-as-evaluator}

The C1 ratio, the C2 path-length / direct-distance metric, the
spatial-diversity metric, and the C6 precision/recall pipeline all
operate on any K direction vectors regardless of how they were
discovered. We use them to evaluate GANSpace, SeFa, LatentCLR,
DisCo, random, and the GT InterFaceGAN/HiGAN boundaries on a single
comparable axis (§\ref{sec:exp-baselines-results}).
\textbf{Methodologically, this re-frames our paper:} we are not a
new direction-discovery method; we are the evaluation framework
through which existing methods can be compared, and through which
genuinely-novel claims about each method's geometric character
(linear vs structural, isotropic vs PCA-like) become measurable.
