\documentclass[10pt]{article} % For LaTeX2e
\usepackage{tmlr}
% If accepted, instead use the following line for the camera-ready submission:
%\usepackage[accepted]{tmlr}
% To de-anonymize for posting to arXiv (preprint servers), instead use:
%\usepackage[preprint]{tmlr}

\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath,amssymb,mathtools}
\usepackage{booktabs}
\usepackage{array}

\newcommand{\R}{\mathbb{R}}
\newcommand{\jvp}{\textsc{jvp}}
\newcommand{\fd}{\textsc{fd}}
\newcommand{\eps}{\varepsilon}

\title{When Finite Differences Flip the Sign of Second-Order\\
Generative Geometry: A Non-Recoverability Result and an Exact Instrument}

% Authors are hidden in the submitted version (anonymous). tmlr.sty suppresses
% this block unless [preprint]/[accepted] is set. Source kept anonymous too so an
% accidental source upload does not de-anonymize. Real metadata for camera-ready:
%   Jihong An <ether9073@gmail.com>, Independent Researcher.
\author{\name Anonymous Author(s) \email anon@example.com \\
      \addr Anonymous Institution}

\def\month{MM}
\def\year{YYYY}
\def\openreview{\url{https://openreview.net/forum?id=XXXX}}

\begin{document}
\maketitle

\begin{abstract}
Studies of the differential geometry of generative models increasingly measure
\emph{second-order} quantities---curvature of the generated manifold and the
terms that enter geodesics. We report a concrete, reproducible hazard in how
these are computed. On Stable Diffusion, a representative second-order geometric
measurement---the rank correlation between the first-order singular value
$\sigma$ of the latent-to-image Jacobian and a second-order curvature ratio
$\rho$ along the leading directions---comes out \emph{with opposite sign}
depending only on the numerical method: the exact forward-mode estimate gives a
consistent strong negative correlation (Spearman $-0.73$, $5/6$ seeds
$\leq\!-0.82$), whereas a finite-difference (\fd) estimate of the same quantity
is sign-unstable and averages $-0.18$. The first-order Jacobian, by contrast, is
recovered by finite differences to $2.4\%$. We explain this asymmetry: central
\fd{} for a \emph{second} derivative is \emph{non-recoverable}---truncation and
floating-point cancellation trade off so that no step size beats a
$\sqrt{\eps_{\mathrm{mach}}}$ relative-error floor, which in the single precision
generators use is large---while \fd{} for a first derivative is well conditioned.
On a StyleGAN generator the naive \fd{} curvature estimate misses the exact
value by $\geq\!45\%$ at every fixed step, in \emph{both} single and double
precision (validated against the exact instrument by the first-order \fd, which
agrees to $2.4\%$); precision does not rescue it. We provide the exact
constant-memory instrument---composed forward-mode automatic differentiation,
generator-agnostic and extending to any order---and survey how the literature
computes second-order geometry, finding that the catastrophic explicit-$/\eps^2$
regime is the natural first thing to reach for but is, fortunately, avoided by
most published methods. Our contribution is the precise account of when finite
differences fail for second-order generative geometry, the practitioner-facing
first/second-order asymmetry, and a drop-in exact replacement.
\end{abstract}

\section{Introduction}
Let $G:\R^{d}\to\R^{3HW}$ be a trained generator. Treating $\{G(z)\}$ as a
manifold pulled back through $G$, a productive line of work studies its geometry:
the pullback metric $M_z=J_z^\top J_z$ and its singular vectors (the local latent
basis), geodesics between latent codes, and the curvature of the generated
manifold \citep{arvanitidis2018,shao2018,chen2018,park2023,saito2025}. First-order
objects ($J_z$, $M_z$) are routine to compute. The \emph{second-order} objects
---curvature, the second-order pushforward $\partial^2 G/\partial\alpha^2$ along a
path, the metric derivative $\partial_z M_z$ that appears in the geodesic
ODE---are where numerical practice matters and, we show, can silently flip a
result.

Our entry point is an observation, not a theorem (Section~\ref{sec:flip}). A
standard second-order measurement on Stable Diffusion returns the
\emph{opposite sign} depending only on whether the second derivative is taken by
finite differences or by exact forward-mode AD. The exact method finds a clean,
seed-consistent relationship; the finite-difference method, applied to the very
same quantity, is sign-unstable and washes it out. A practitioner choosing the
natural finite-difference route would draw the wrong qualitative conclusion.

The explanation is an asymmetry that is elementary once stated but, in our
experience, not widely appreciated in this applied setting
(Section~\ref{sec:fd}): a central finite difference is well conditioned for a
\emph{first} derivative but \emph{non-recoverable} for a \emph{second}
derivative, because for the second derivative truncation error and floating-point
cancellation cannot be made small at the same step size, leaving a relative-error
floor of order $\sqrt{\eps_{\mathrm{mach}}}$. In the single precision that GANs
and diffusion U-Nets run in, that floor is large; and on a strongly nonlinear,
non-smooth generator it is large at \emph{every} step and in \emph{both}
precisions (Section~\ref{sec:stylegan}: $\geq\!45\%$ on StyleGAN, fp32 and fp64
alike).

We make three contributions: (i) the reproducible sign-inversion phenomenon and
the first/second-order asymmetry that governs it, as concrete guidance; (ii) an
exact, constant-memory instrument for the higher-order pushforward via composed
forward-mode AD, which removes the hazard at the cost of one extra forward pass
per order; and (iii) a survey of how the field computes second-order geometry,
locating where the trap is and is not. We are explicit about scope
(Section~\ref{sec:scope}): we do not claim any published result is wrong---indeed
we find most published methods route around the catastrophic regime---we
characterize a real numerical hazard and supply the fix.

\section{The sign-inversion phenomenon}
\label{sec:flip}
On a Stable-Diffusion h-space pipeline we take a base latent, form the first-order
Jacobian $J$ of the (truncated) denoising map along $K$ random h-space probes,
and along its top-$k$ right-singular directions $u_i$ measure a second-order
curvature ratio
\[
\rho_i \;=\; \frac{|\,\partial^2_\alpha G(z+\alpha u_i)\,|}{|\,\partial_\alpha
G(z+\alpha u_i)\,|}\Bigg|_{\alpha=0},
\]
the magnitude of the second-order pushforward relative to the first. We then ask a
standard question---does curvature co-vary with the first-order singular value
$\sigma_i$?---via Spearman$(\sigma,\rho)$, computing the second derivative in
$\rho$ two ways: exact composed \jvp{} (Section~\ref{sec:method}) and a finite
difference of the \jvp{} (the natural way a practitioner would estimate it).

Over $6$ seeds (top-$12$ directions each), the two methods disagree in
\emph{sign} (Table~\ref{tab:flip}). The exact instrument yields a strong negative
correlation on five of six seeds (mean $-0.73$; the five clustered in
$[-0.94,-0.82]$, with seed~4 a near-zero outlier at $-0.03$). The
finite-difference estimate of the \emph{same} ratio is sign-unstable across seeds
and averages $-0.18$: it does not merely add noise, it destroys the relationship.
The exact estimate is more negative than the \fd{} estimate on \emph{all six}
seeds (Wilcoxon signed-rank $p=0.031$, the floor attainable at $n{=}6$), and the
\fd{} estimate is sign-unstable at \emph{every} step in the grid
$\{0.1,0.05,0.01,0.001\}$ (means between $-0.08$ and $-0.34$): no step reproduces
the clean $-0.73$, so the disagreement is not an artifact of the chosen step. A study
relying on the finite-difference route would report ``no clear relationship,'' the
opposite of what the exact second-order estimate finds.

\begin{table}[h]\centering
\begin{tabular}{lccccccc}
\toprule
seed & 0 & 1 & 2 & 3 & 4 & 5 & \textbf{mean} \\
\midrule
exact composed \jvp & $-0.85$ & $-0.83$ & $-0.92$ & $-0.94$ & $-0.03$ & $-0.84$ & $\mathbf{-0.73}$ \\
\fd-of-\jvp ($\eps{=}0.05$) & $+0.18$ & $-0.55$ & $-0.78$ & $-0.40$ & $+0.55$ & $-0.07$ & $\mathbf{-0.18}$ \\
\bottomrule
\end{tabular}
\caption{Spearman$(\sigma,\rho)$ on Stable Diffusion, per seed. The exact
second-order estimate is consistently strongly negative; the finite-difference
estimate of the identical quantity changes sign across seeds and averages near
zero. ($\rho$ is a measurement we define to expose the method dependence; the
point is that the two numerical routes disagree on its sign, not the specific
value of any prior work.)}
\label{tab:flip}
\end{table}

\section{Why: finite differences are non-recoverable for a second derivative}
\label{sec:fd}
For smooth scalar $f$ the central second difference is
$D^2_\eps f=\big(f(\alpha+\eps)-2f(\alpha)+f(\alpha-\eps)\big)/\eps^2$, with
\[
D^2_\eps f - f''(\alpha)
=\underbrace{\tfrac{\eps^2}{12}f^{(4)}(\alpha)+O(\eps^4)}_{\text{truncation}}
+\underbrace{O\!\big(\eps_{\mathrm{mach}}\,|f|/\eps^2\big)}_{\text{cancellation}} .
\]
The two terms cannot be made small together: minimizing their sum gives an optimal
step $\eps^\star\sim\eps_{\mathrm{mach}}^{1/4}$ and a relative-error floor (dividing
the absolute error above by $|f''|$, assuming $f^{(4)}$ bounded, $f''\neq0$, and
$|f|$ normalized to $O(1)$) of $2\sqrt{|f^{(4)}f|}/|f''|\cdot\eps_{\mathrm{mach}}^{1/2}$.
The \emph{exponent} $\tfrac12$ is intrinsic to the operation; the constant carries
the generator's nonlinearity $|f^{(4)}f|/f''^2$, which is large for a deep network.
In double precision the floor is $\sim\!10^{-8}$ in the benign (constant~$\approx1$)
case; in the single precision generators use
($\eps_{\mathrm{mach}}\!\approx\!1.2\times10^{-7}$ for fp32) it is
$\sim\!3\times10^{-4}$ benign and far worse for a strongly nonlinear $G$
(Section~\ref{sec:stylegan}: $\geq\!45\%$). No choice of $\eps$ removes the floor. Because the floor is a property of the
finite-difference \emph{operation} in fixed precision, not of any architecture, the
hazard is general; we demonstrate its magnitude on two generator families
(StyleGAN and a latent-diffusion U-Net).

\paragraph{The asymmetry.} The \emph{first} central difference
$\big(f(\alpha+\eps)-f(\alpha-\eps)\big)/2\eps$ has truncation $O(\eps^2)$ and
cancellation $O(\eps_{\mathrm{mach}}|f|/\eps)$, giving a much milder floor
$\sim\eps_{\mathrm{mach}}^{2/3}$ and a wide usable window. This is why first-order
Jacobians are safely finite-differenced while second-order curvature is not---and
it is the control we use to validate our instrument (Sections~\ref{sec:flip},
\ref{sec:stylegan}): wherever the first-order \fd{} agrees with exact \jvp, the
pipeline is correct and any second-order disagreement is genuine.

\paragraph{Analytic referee.} Let
$f(\alpha)=[\,e^{0.7\alpha},\sin 3\alpha,(\alpha+0.4)^3,\tanh 2\alpha,
\alpha^2\cos\alpha\,]$, with closed-form derivatives. At $\alpha=0.3$ in double
precision (Table~\ref{tab:toy}) the central-\fd{} second-order relative error is
U-shaped, bottoming at $8.6\times10^{-9}$ near $\eps=10^{-4}$ and diverging to
$0.29$ by $\eps=10^{-8}$; the composed \jvp{} attains $2.5\times10^{-17}$---\fd's
best is $3.5\times10^{8}$ times worse.

\begin{table}[h]\centering
\begin{tabular}{rcl}
\toprule
$\eps$ & central \fd{} 2nd-order rel.\ err. & \\
\midrule
$10^{-1}$ & $7.9\times10^{-3}$ & (truncation) \\
$10^{-2}$ & $8.0\times10^{-5}$ & \\
$10^{-4}$ & $8.6\times10^{-9}$ & (best) \\
$10^{-6}$ & $3.0\times10^{-5}$ & \\
$10^{-8}$ & $2.9\times10^{-1}$ & (cancellation) \\
\midrule
composed \jvp & $2.5\times10^{-17}$ & (exact) \\
\bottomrule
\end{tabular}
\caption{Second-order accuracy on the analytic toy at $\alpha=0.3$, double
precision. Finite differences never approach the exact composed \jvp.}
\label{tab:toy}
\end{table}

\section{Quantification on a real generator}
\label{sec:stylegan}
The toy fixes ideas; on a real generator the failure is both worse and broader
than cancellation alone. On a StyleGAN-bedroom generator ($256{\times}256$) we
form the naive \fd{} curvature estimate a practitioner would write,
$D^2_\delta G=\big(G(z{+}\delta u)-2G(z)+G(z{-}\delta u)\big)/\delta^2$, and the
exact composed \jvp{} $\partial^2_\alpha G(z{+}\alpha u)$, over $6$ directions
($3$ attributes, $3$ random) $\times\,16$ samples, sweeping $\delta$
(Table~\ref{tab:stylegan}). The exact curvature is a definite, step-free number
(mean $1.17$). Holding a \emph{single} step across all samples---what a
practitioner does---the \fd{} mean relative error never drops below $45.2\%$ (best
step $\delta=3$, bracketed on both sides); it is $267\%$ at $\delta\!\sim\!0.3$ and
explodes as $\delta\to0$. Even allowing an \emph{oracle} per-sample step the median
best error is still $\sim\!9\%$ (vs.\ the composed \jvp's $0$). \textbf{Precision
does not rescue it:} rerunning the entire sweep in fp64 leaves the floor unchanged
($47.1\%$ in both fp32 and fp64; Table~\ref{tab:stylegan} note). The mechanism is
not cancellation alone: at usable steps the error is truncation (precision-blind);
at tiny steps fp32 adds catastrophic cancellation, but fp64---which removes
cancellation---instead exposes the generator's \emph{non-smoothness} (LeakyReLU
kinks), where the chord-scale second difference does not converge to the pointwise
pushforward (fp64 error grows to $>\!3000\%$ as $\delta\to0$). So on a real
generator \emph{no step and no precision} recovers the exact second-order
pushforward, which the composed \jvp{} returns directly. The control: the
\emph{first}-order central difference of the same pipeline agrees with the exact
first-order \jvp{} to $2.4\%$, so the instrument and pipeline are correct and the
second-order failure is genuine, not a bug.

\begin{table}[h]\centering
\begin{tabular}{rccc}
\toprule
step $\delta$ & $|D^2_\delta G|$ & 2nd-order rel.\ err. & 1st-order rel.\ err. \\
\midrule
$8$ & $7.1\times10^{-3}$ & $74.5\%$ & $63.8\%$ \\
$3$ & $2.7\times10^{-2}$ & $\mathbf{45.2\%}$ (floor) & $44.1\%$ \\
$1$ & $9.9\times10^{-2}$ & $106.4\%$ & $26.2\%$ \\
$0.3$ & $3.7\times10^{-1}$ & $266.9\%$ & $12.5\%$ \\
$0.1$ & $1.06$ & $622.8\%$ & $5.5\%$ \\
$0.02$ & $4.05$ & $6.8\times10^{3}\%$ & $\mathbf{2.4\%}$ (1st-order min) \\
$10^{-3}$ & $4.09\times10^{2}$ & $2.0\times10^{6}\%$ & $139\%$ \\
\midrule
composed \jvp & $1.17$ & exact (step-free) & --- \\
\bottomrule
\end{tabular}
\caption{StyleGAN-bedroom, fp32. Columns are means over $6$ directions $\times16$
samples at each fixed step; rel.\ err.\ is the mean per-direction relative error,
while $|D^2_\delta G|$ is the aggregate magnitude (not its direct source). The
naive \fd{} second difference never gets within $45\%$ of the exact curvature at
any fixed step; the first-order \fd{} of the same code agrees to $2.4\%$,
validating the instrument. Rerunning in fp64 leaves the best-step floor unchanged
($47.1\%$ fp32 vs.\ $47.1\%$ fp64), so the floor is not a single-precision
artifact---it is truncation at usable steps and generator non-smoothness at tiny
steps.}
\label{tab:stylegan}
\end{table}

\section{The exact instrument: composed forward-mode AD}
\label{sec:method}
Forward-mode AD evaluates, for a map $f$ and tangent $v$, the
Jacobian--vector product $\partial f(x)[v]$ alongside $f(x)$ in one pass without
forming the Jacobian. With $f=G$, $x=\gamma(\alpha)$, $v=u$ this gives the exact
first-order pushforward. The second-order term is the directional derivative of
the first, so composing forward mode with itself,
$\partial^2_\alpha G(\gamma(\alpha))=\partial_\alpha\big[\partial
G(\gamma(\alpha))[u]\big][u]$, is exact and is evaluated by nesting two \jvp{}
calls (\texttt{torch.func.jvp} of a function that returns a \jvp). No Jacobian is
stored, so memory stays at one forward pass; the same nesting yields any order in
a constant number of passes. The instrument is generator-agnostic (we ran it
unmodified on StyleGAN1/2 and a latent-diffusion U-Net) and, unlike reverse-mode
Hessian materialization (memory $O(3HW)$), its cost is independent of output
dimension.

\section{Survey: where the trap is, and is not}
\label{sec:survey}
Table~\ref{tab:survey} classifies representative latent-geometry papers by how
they obtain second-order/geodesic quantities. The explicit $/\eps^2$ second-order
finite difference---the catastrophic regime of Section~\ref{sec:fd}---is the
natural first thing to reach for, but most published methods avoid it: they keep
the metric first-order, replace the geodesic ODE with a discretized energy whose
\emph{exact} gradient is a well-conditioned discrete Laplacian (not a $/\eps^2$
curvature), use analytic Christoffel symbols, or differentiate the energy with
autodiff \citep{yang2018}; first-order constructions
\citep{park2023,chen2018,kwon2023} never touch a second derivative. The genuinely
exposed practice is first-order \fd{} of a network derivative (e.g.\ a
score-Jacobian--vector product), which is only mildly inaccurate, and---most
relevant here---the direct $/\eps^2$ curvature estimate that a practitioner adds
on top of an existing pipeline, as in our Section~\ref{sec:flip} measurement. The
contribution of the survey is to draw this line precisely. The reason the field is
mostly safe is structural---the canonical formulations are first-order metrics or
energy minimizations \emph{by construction}, which never form an explicit
second-order finite difference. But that safety is a property of the current
formulations, not a law: as curvature- and geodesic-level analysis of diffusion
models grows (e.g.\ \citet{saito2025}, and any second-order extension of the
first-order bases of \citet{park2023}), the $/\eps^2$ estimate becomes the obvious
next step. We intend the line in Table~\ref{tab:survey}, the asymmetry, and the
drop-in instrument as forward-looking guardrails for that emerging work, not as a
retrospective indictment.

\begin{table}[h]\centering
\begin{tabular}{>{\raggedright}p{4.4cm} l p{3.6cm}}
\toprule
Paper & 2nd-order quantity & how obtained \\
\midrule
\citet{shao2018} & discretized geodesic energy &
exact discrete-energy gradient (Laplacian), coarse $T$; \emph{no} $/\eps^2$ \\
\citet{saito2025} & geodesic (energy) & first-order score-differences $\|s(x_{i+1}){-}s(x_i)\|^2$ (no $/\eps^2$); JVP backend unspecified \\
\citet{arvanitidis2018} & geodesic ODE (Christoffel) & analytic $\partial M$, numeric ODE integ. \\
\citet{yang2018} & geodesic energy & autodiff \\
\citet{park2023} & none (first-order basis) & autodiff power iteration \\
\citet{chen2018} & none (first-order metric) & --- \\
\citet{kwon2023} & none (first-order) & CLIP-grad + MLP \\
\bottomrule
\end{tabular}
\caption{How representative latent-geometry papers obtain second-order quantities.
The explicit $/\eps^2$ second-order finite difference (Sections~\ref{sec:fd},
\ref{sec:stylegan}) is largely avoided in published work; it is nonetheless the
natural choice for a practitioner adding a curvature measurement, which is where
the hazard bites.}
\label{tab:survey}
\end{table}

\paragraph{Guidance.} The practical rule is short. \emph{First}-order quantities
(Jacobians, pullback metrics, first-order pushforwards) may be finite-differenced;
choose a step near $\eps_{\mathrm{mach}}^{1/3}$ and expect a few digits. \emph{Second
or higher}-order quantities (curvature, the second-order pushforward, the metric
derivative / Christoffel symbols, anything with a $/\eps^2$) should \emph{not} be
finite-differenced: in single precision no step gives even one reliable digit on a
deep generator (Section~\ref{sec:stylegan}), and the error can flip the sign of a
downstream conclusion (Section~\ref{sec:flip}). Use composed forward-mode AD instead
---it is exact, costs one extra forward pass per order, and adds no memory. A free
diagnostic comes with it: compare the first-order finite difference to the
first-order \jvp; if they disagree, the pipeline is wrong, and if they agree, any
second-order disagreement is real, not numerical.

\section{Scope, limitations, and what we do not claim}
\label{sec:scope}
This is a numerical-methods contribution with a survey and demonstrations.
\textbf{(1)} We do \emph{not} claim any surveyed paper reached a false conclusion;
on the contrary, the survey shows most route around the catastrophic regime. The
sign inversion of Section~\ref{sec:flip} is on a quantity \emph{we} define to
expose method dependence, not a number from prior work.
\textbf{(2)} The non-recoverability result is specifically about \emph{second}-order
finite differences; first-order finite differences (e.g.\ score-JVPs) are well
conditioned, and we say so.
\textbf{(3)} The instrument is a thin, exact wrapper around forward-mode AD; its
value is in routine exact higher-order pushforward and in the precise account of
when the finite-difference alternative fails, not in algorithmic novelty.
\textbf{(4)} We do not propose a downstream task win: this is a
measurement-correctness result, not a demonstration that exact curvature improves
an application. Whether accurate second-order geometry changes downstream
conclusions is left to future work.

\section{Reproducibility}
We release the exact-instrument module, the analytic referee, and the StyleGAN
and Stable-Diffusion scripts with fixed seeds and the emitted
\texttt{metrics.json} for every table. All experiments run on a single 8\,GB GPU.

\bibliography{refs}
\bibliographystyle{tmlr}

\end{document}
